---
title: 'Statistical Inference Course Project Part 1 : Central Limit Theorem using
  Simulation'
author: "Jonathan Lok-Chuen Lo"
date: "October 25, 2015"
output: pdf_document
geometry: margin=0.75in
---

```{r set_options, echo=FALSE}
knitr::opts_chunk$set(
    echo = TRUE, 
    warning = FALSE, 
    message = FALSE,
    fig.path = "figure/",
    fig.height = 4,
    fig.width = 6,
    fig.pos = "center")
```

# Overview

In this project, I will investigate the exponential distribution in R and 
compare it with the Central Limit Theorem. The exponential distribution is used
and simulated in R with `rexp(n, lambda)` where lambda is the rate parameter. 
I will investigate the distribution of 40 exponentials and set lambda = 0.2.
The simulated mean and standard deviation will be compared against the
theoretical values to illustrate the Central Limit Theorem.

# Simulations

Generate the simulated data from exponential distribution with lambda = 0.2 and 
number of exponentials = 40. A thousand simulations is produced, and hence total
of 40,000 simulated data is required.

```{r simulate_data}
# Initialise the seed to ensure reproducibility
set.seed(8)

# Parameters for the simulation
n <- 40
lambda <- 0.2
numSimulations <- 1000

# Generate the simulated data and compare the sample mean and variance
simulatedData <- matrix(rexp(n * numSimulations, lambda), numSimulations, n)
simulatedDataMean <- apply(simulatedData, 1, mean)
simulatedDataVar <- apply(simulatedData, 1, var)
```

# Sample Mean versus Theoretical Mean

Calculate the average sample mean and compare it against the theoretical mean of
exponential distribution, which is equal to 1/lambda.

```{r compute_mean}
sampleMean <- mean(simulatedDataMean)
theoreticalMean <- 1 / lambda

# Print the results
cbind(sampleMean, theoreticalMean)
```

Tha sample mean is **`r round(sampleMean, 4)`** while the theoretical 
mean is **`r round(theoreticalMean, 4)`**. 

Following is a plot of the distribution of the sample mean distribution. This 
shows that the sample mean is a very good estimate of the theoretical mean. The 
corresponding normal distribution is plotted in blue as a reference and the 
sample mean does appear to be close to normal.

```{r plot_mean, echo=FALSE}
h <- hist(
    simulatedDataMean,
    main = "Distribution of the Sample Mean of the Simulated Data",
    xlab = "Sample Mean",
    col = "gray",
    cex.lab = 0.75)

abline(v = sampleMean, col = "red", lwd = 2)
abline(v = theoreticalMean, col = "green", lwd = 2)

legend(
    "topright", 
    legend = c("AVerage Sample Mean", "Theoretical Mean"),
    col = c("red", "green"),
    lwd = 2,
    cex = 0.75)

xfit <- seq(from = 0, to = 8, length.out = 100) 
yfit <- dnorm(xfit, mean = mean(simulatedDataMean), sd = sd(simulatedDataMean))
yfit <- yfit * diff(h$mids[1:2]) * length(simulatedDataMean) 

lines(xfit, yfit, col = "blue", lwd = 2)
```

# Sample Variance versus Theoretical Variance

Calculate the average sample variance and compare it against the theoretical 
variance of exponential distribution, which is equal to (1/lambda)^2.

```{r compute_var}
sampleVar <- mean(simulatedDataVar)
theoreticalVar <- (1 / lambda)^2

# Print the results
cbind(sampleVar, theoreticalVar)
```

Tha sample variance is **`r round(sampleVar, 4)`** while the theoretical 
variance is **`r round(theoreticalVar, 4)`**. 

Following is a plot of the distribution of the sample variance. This again shows
that the sample variance is a very good estimate of the theoretical variance. 
The corresponding normal distribution of the sample variance is plotted in blue 
and the sample variance does appear to be close to normal.

```{r plot_var, echo=FALSE}
h <- hist(
    simulatedDataVar,
    main = "Distribution of the Sample Variance of the Simulated Data",
    xlab = "Sample Variance",
    col = "gray",
    cex.lab = 0.75)

abline(v = sampleVar, col = "red", lwd = 2)
abline(v = theoreticalVar, col = "green", lwd = 2)

legend(
    "topright", 
    legend = c("Average Sample Variance", "Theoretical Variance"),
    col = c("red", "green"),
    lwd = 2,
    cex = 0.75)

xfit <- seq(from = 0, to = 85, length.out = 100) 
yfit <- dnorm(xfit, mean = mean(simulatedDataVar), sd = sd(simulatedDataVar))
yfit <- yfit * diff(h$mids[1:2]) * length(simulatedDataVar) 

lines(xfit, yfit, col = "blue", lwd = 2)
```


# Distribution

Below is a plot of the exponential distribution with number of exponentials = 40
and lambda = 0.2. It can be seen that although the distribution is very 
different to a normal distribution, the distribution of the sample mean and 
sample variance are close to normal as shown earlier.

```{r plot_dist, echo=FALSE}
hist(
    simulatedData,
    main = "Exponential Distribution with n = 40 and lambda = 0.2",
    xlab = "Data",
    col = "gray",
    cex.lab = 0.75)

abline(v = 1/lambda, col = "red", lwd = 2)

legend(
    "topright", 
    legend = c("Theoretical Mean"),
    col = c("red"),
    lwd = 2,
    cex = 0.75)
```


